Variance reduction for MCMC methods via martingale representations

### 2020-03-10 aap/Revision/vr-mcmc-revision-new.tex
theorem 4 update

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{q,l})_{l=0}^q\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
\begin{eqnarray}
\label{eq:qpl}
Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
\end{eqnarray}
In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
\end{equation}
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the time homogeneous case ($\Phi_l = \Phi$ for all $l \in \nset$) and the sample average
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
\end{equation}
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{eqnarray}
\nonumber
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X^x_{l-1})\phi_k(\xi_{l})\right]
\\
&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X^x_{l-1})\phi_{k}\left(\xi_{l}\right),
\label{eq:29032018a5}
\end{eqnarray}
where
\begin{equation}
\label{eq:definition-A-s,k}
A_{s,k}(x)=\frac{1}{s}\sum_{r=1}^{s}\bar a_{r,k}(x).
\end{equation}
Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
\[
\PE[A_{N+n-l,k}(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[A_{N+n-l,k}(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
\]
and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\begin{equation}
\label{eq:expression-variance}
\PVar[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})]
\end{equation}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. We use the notations of example \ref{exam:langevin-algorithm}. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.

By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=1}^{n}\Bigl(1-\frac{l}{n}\Bigr)A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
\label{eq:29032018a5}
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i.\)
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{equation}
\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
\end{equation}
Our analysis is carried out under the following two assumptions:
\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L < \infty$ such that
\begin{eqnarray*}
\| \nabla U(x)-\nabla U(y) \| \leq L \|x-y\|, \quad x,y\in \rset^d.
\end{eqnarray*}
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0$ and $m>0$ such that for any $x\not\in B_{K}(0)$  it holds
\begin{equation*}
\ps{D^2U(x)}{x} \geq m\|x\|^2.
\end{equation*}
\end{description}
Let $\pi$ be the probability measure on $\rset^d$ with density $\pi(x)$ of the form \eqref{eq:stationary_distr};
for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one step of the ULA algorithm by
\begin{equation}
\label{eq:ula_kernel}
\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y
\end{equation}
for any $A \in \mathcal{B}(\rset^d)$. Note that for a homogeneous Markov chain $\left(X^x_{p}\right)_{p \in \nset_0}$ of the form \eqref{eq:chain} with constant step size $\gamma$, $\mathsf{P}(X_n^x \in A) = \kerULA^n_{\gamma}(x,A)$.

@@

\begin{thm}\label{th:mr}
Assume {\bf (H1)} and {\bf (H2)}.
Suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K \times d \geq 2$ times continuously differentiable
and for all $x\in\mathbb R^d$ and  \(\mathbf{k}\) satisfying \(0<\|\mathbf{k}\|\leq  K,\)
\begin{equation}
\label{eq:smooth-mu}
|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad |\partial^{\mathbf{k}} \mu (x)|\leq B_\mu \eqsp.
\end{equation}
Then it holds
\begin{eqnarray}
\label{eq:var-bound}
\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-2} \eqsp,
\end{eqnarray}
%where \tcr{\(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)}
\end{thm}
Let us sketch the main steps of the proof. First using integration by parts we prove that
\begin{eqnarray}
\label{eq:Ask-part}
A_{s,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_s) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right]
\end{eqnarray}
where $F_s(x,Z_1,\ldots,Z_s)=s^{-1}\sum_{r=0}^s f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part}, we derive
\begin{eqnarray*}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1} A^2_{s,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma}{2}\right)^{|I|K}
\PVar \left( s^{-1} \sum\nolimits_{p=1}^{s}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
\right),
\end{eqnarray*}
where the sum runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}$ and
\[
\mathbf{\mathbf{K}}_{I}=K(\indi{I}(1),\ldots, \indi{I}(d)).
\]
 Finally we show via  the Poincare inequality that under smoothness assumption \eqref{eq:smooth-mu} it holds
 \eric{Since we do not state was $W_s$ is, it is a bit difficult to guess why we are using a Poincare inequality there}
\begin{equation*}
\PVar \left(s^{-1} \sum\nolimits_{p=1}^{s}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
\right)\leq W_s(x)
\end{equation*}
for all $x$ and some family of functions $(W_s)_{s\geq 1}$ such that  the sum $\sum_{s=1}^n \mathsf{E}[W_s(X_{N+n-s-1})]$ is bounded uniformly in $n\in \mathbb{N}.$
\eric{Speak about truncation here.}

\section{Numerical results}
\label{sec:num}
We perform numerical experiments based on ULA algorithm \ref{exam:langevin-algorithm} with constant step size $\gamma$. We construct a polynomial approximation  for each \(Q_r(x)\) in \eqref{eq:definition-bar-a} of the form:
\begin{equation*}
\widehat{Q}_{r}(x) = \sum_{\|\mathbf{s}\|\leq m} \widehat\beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d) \in \nset_0^d
\end{equation*}
with some coefficients \(\widehat \beta_{\mathbf{s}}\in \mathbb{R}\) obtained using a modified least-squares criteria based on $T$ independent training trajectories \(\left\{\left(X_1^{(s)},\ldots,X_{N+n}^{(s)}\right)\right\}_{s=1}^{T}\) with $N$ being the size of burn-in period:
\begin{eqnarray}
\label{eq:Qregr}
\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
\end{eqnarray}
for \(1 \leq r \leq n-1,\) $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition.
Then using the identity
\[
\xi^j = j! \sum_{r = 0}^{\lfloor j/2 \rfloor} \frac{1}{2^r  r! \sqrt{(j-2r)!}} H_{j-2r}(\xi),\quad \xi \in \mathbb{R},
\]
we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ in \eqref{eq:definition-bar-a}. Namely, for all $x \in \mathbb{R}^d$,
\begin{eqnarray*}
\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet \vee \mathcal{G}_N}
= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{i,k_i,s_i}(x),
\end{eqnarray*}
where for all integers $i,k_i,s_i$ and $x \in \rset^d$,
\begin{align*}
P_{r,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x - \gamma \mu(x) + \sqrt{\gamma} \xi)^{s} \right]
\end{align*}
is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients. In consequence, it is enough to estimate functions \(Q_{r}(x)\). It can be done using a modified least-squares criteria based on $T$ training trajectories \(\left(X_1^{(s)},\ldots,X_{N+n}^{(s)}\right), s = 1,\ldots,T\) with $N$ being the size of burn-in period:
\begin{eqnarray}
\label{eq:Qregr}
\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
\end{eqnarray}
for \(1 \leq r \leq n-1,\) $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition. Certainly one may use another functional class $\Psi$ in $\label{eq:Qregr}$, but using polynomials allows to compute closed-form representation of $\widehat a_{r,\mathbf{k}} (x)$.
\par
Due to Lemma~\ref{a_decay},
\sergey{we need to put some lemma about the truncation point here, I will do it}
it is enough to estimate $Q_{r}$ for $r < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}(x)$.

@@

\PE\left[
s^{-1} \sum\nolimits_{r=1}^{s}\{\partial_{1}^{\mathbf{k}'} f(X^x_r)-\PE[\partial_{1}^{\mathbf{k}'}f(X^x_r)]\}\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_1)\right]

@@

\left\{\PE\left[\frac{1}{s}\sum_{p=1}^{s}\left\{\partial_{1}^{\mathbf r}f\left(X^x_{p}\right)-\PE\left[\partial_{1}^{\mathbf r}f\left(X^x_{p}\right)\right]\right\}\mathbf{H}_{\mathbf{q}}(Z_{1})\right]\right\}^{2}.

@@

\PVar\left(s^{-1}\sum\nolimits_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)

@@

\PVar\left(s^{-1}\sum\nolimits_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)
\right)\leq W_s(x)

@@

for all $x$ and some family of functions $(W_s)_{s\geq 1}$ such that  the sum $\sum_{s=1}^n \mathsf{E}[W_s(X_{N+n-s-1})]$ is bounded uniformly in $n\in \mathbb{N}.$

>>>>

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{q,l})_{l=0}^q\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
\begin{eqnarray}
\label{eq:qpl}
Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
\end{eqnarray}
In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
\end{equation}
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the time homogeneous case ($\Phi_l = \Phi$ for all $l \in \nset$) and the sample average
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
\end{equation}
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{eqnarray}
\nonumber
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X^x_{l-1})\phi_k(\xi_{l})\right]
\\
&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
\label{eq:29032018a5}
\end{eqnarray}
Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
\[
\PE[g(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[g(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
\]
for any measurable function $g$  and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\begin{equation}
\label{eq:expression-variance}
\PVar_x[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\frac{1}{n^2}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
\end{equation}
where 
\begin{eqnarray}
\label{eq:definition-A-s,k}
A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N}
\end{eqnarray}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. We use the notations of example \ref{exam:langevin-algorithm}. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.

By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\frac{1}{n}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
\label{eq:29032018a5}
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i.\)
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{equation}
\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
\end{equation}
Our analysis is carried out under the following two assumptions:
\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L < \infty$ such that
\begin{eqnarray*}
\| \nabla U(x)-\nabla U(y) \| \leq L \|x-y\|, \quad x,y\in \rset^d.
\end{eqnarray*}
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0$ and $m>0$ such that for any $x\not\in B_{K}(0)$  it holds
\begin{equation*}
\ps{D^2U(x)}{x} \geq m\|x\|^2.
\end{equation*}
\end{description}
Let $\pi$ be the probability measure on $\rset^d$ with density $\pi(x)$ of the form \eqref{eq:stationary_distr};
for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one step of the ULA algorithm by
\begin{equation}
\label{eq:ula_kernel}
\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y
\end{equation}
for any $A \in \mathcal{B}(\rset^d)$. Note that for a homogeneous Markov chain $\left(X^x_{p}\right)_{p \in \nset_0}$ of the form \eqref{eq:chain} with constant step size $\gamma$, $\mathsf{P}(X_n^x \in A) = \kerULA^n_{\gamma}(x,A)$.

@@

\begin{thm}\label{th:mr}
Assume {\bf (H1)} and {\bf (H2)}.
Suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K \times d \geq 2$ times continuously differentiable
and for all $x\in\mathbb R^d$ and  \(\mathbf{k}\) satisfying \(0<\|\mathbf{k}\|\leq  K,\)
\label{eq:smooth-mu}
|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad |\partial^{\mathbf{k}} \mu (x)|\leq B_\mu \eqsp.
Then  it holds
\label{eq:var-bound}
\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-1} \eqsp,
provided $n\gamma^2$ is bounded.
%where \tcr{\(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)}
\end{thm}
Let us sketch the main steps of the proof. First using integration by parts we prove that
\label{eq:Ask-part}
A_{s,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_s) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right]
where $F_s(x,Z_1,\ldots,Z_s)=\sum_{r=0}^s f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part}, we derive
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1} A^2_{s,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\PVar_x \left( \sum_{p=1}^{s}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
\right),
where the sum runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}$ and
\mathbf{\mathbf{K}}_{I}=K(\indi{I}(1),\ldots, \indi{I}(d)).
 Finally, we show  that under our smoothness assumption \eqref{eq:smooth-mu} it holds
\PVar \left(\sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
\right)\leq C(1+V(x))
for some constant $C$ not depending on $q$ and $V(x) = 1 + \|x\|^2$.
\section{Numerical results}
\label{sec:num}
We perform numerical experiments based on ULA algorithm \ref{exam:langevin-algorithm} with constant step size $\gamma$. We construct a polynomial approximation  for each \(Q_r(x)\) in \eqref{eq:definition-bar-a} of the form:
\widehat{Q}_{r}(x) = \sum_{\|\mathbf{s}\|\leq m} \widehat\beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d) \in \nset_0^d
with some coefficients \(\widehat \beta_{\mathbf{s}}\in \mathbb{R}\) obtained using a modified least-squares criteria based on $T$ independent training trajectories \(\left\{\left(X_1^{(s)},\ldots,X_{N+n}^{(s)}\right)\right\}_{s=1}^{T}\) with $N$ being the size of burn-in period:
\label{eq:Qregr}
\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
for \(1 \leq r \leq n-1,\) $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition.
Then using the identity
\xi^j = j! \sum_{r = 0}^{\lfloor j/2 \rfloor} \frac{1}{2^r  r! \sqrt{(j-2r)!}} H_{j-2r}(\xi),\quad \xi \in \mathbb{R},
we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ in \eqref{eq:definition-bar-a}. Namely, for all $x \in \mathbb{R}^d$,
\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet \vee \mathcal{G}_N}
= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{i,k_i,s_i}(x),
where for all integers $i,k_i,s_i$ and $x \in \rset^d$,
P_{r,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x - \gamma \mu(x) + \sqrt{\gamma} \xi)^{s} \right]
is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients. In consequence, it is enough to estimate functions \(Q_{r}(x)\). It can be done using a modified least-squares criteria based on $T$ training trajectories \(\left(X_1^{(s)},\ldots,X_{N+n}^{(s)}\right), s = 1,\ldots,T\) with $N$ being the size of burn-in period:
\label{eq:Qregr}
\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
for \(1 \leq r \leq n-1,\) $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition. Certainly one may use another functional class $\Psi$ in $\label{eq:Qregr}$, but using polynomials allows to compute closed-form representation of $\widehat a_{r,\mathbf{k}} (x)$.
\par
Due to Lemma~\ref{a_decay},
\sergey{we need to put some lemma about the truncation point here, I will do it}
it is enough to estimate $Q_{r}$ for $r < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}(x)$.

@@

\PE\left[ \sum\nolimits_{r=1}^{s}\{\partial_{1}^{\mathbf{k}'} f(X^x_r)-\PE[\partial_{1}^{\mathbf{k}'}f(X^x_r)]\}\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_1)\right]

@@

\left\{\PE\left[\sum_{p=1}^{s}\left\{\partial_{1}^{\mathbf r}f\left(X^x_{p}\right)-\PE\left[\partial_{1}^{\mathbf r}f\left(X^x_{p}\right)\right]\right\}\mathbf{H}_{\mathbf{q}}(Z_{1})\right]\right\}^{2}.

@@

\PVar\left(\sum_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)

@@

\PVar\left(\sum_{p=1}^{q}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)
\right)\leq C(1+V(x))

@@

for all $x$ and some constant $C$  not depending on $q.$
To keep the notational burden at a reasonable level, we present the proof only in one-dimensional case.
Multidimensional extension is straightforward but requires involved notations.
First, we need to prove several auxiliary results.
\begin{lem}\label{lem:06062018a1}
Let $(x_p)_{p\in\mathbb N_0}$
and $(\epsilon_p)_{p\in\mathbb N}$
be sequences of nonnegative real numbers
satisfying $x_0=\ol C_0$ and
0&\le x_p\le\alpha_p x_{p-1}+\gamma \epsilon_p,\quad p\in\mathbb N,
\label{eq:06062018a1}\\
0&\le\epsilon_p\le\ol C_1\prod_{k=1}^p \alpha_k^2,\quad p\in\mathbb N,
\label{eq:06062018a2}
where $\alpha_p,\gamma\in(0,1)$, $p\in\mathbb N$,
and $\ol C_0,\ol C_1$ are some nonnegative constants. Assume
\begin{equation}\label{eq:06062018a3}
\gamma \sum_{r=1}^\infty \prod_{k=1}^r \alpha_k\le\ol C_2
for some constant $\ol C_2$. Then
x_p\le(\ol C_0+\ol C_1\ol C_2)\prod_{k=1}^p \alpha_k,\quad p\in\mathbb N.
\end{lem}
